---
title: "Getting started"
output:
  html_notebook:
    theme: simplex
  html_document:
    df_print: paged
---

You can execute R scripts in three ways:

+ from the Command line (R command)
+ from the R gui (R.app on the mac, offers a simple development environment)
+ from the development environment of Rstudio. 

We will focus on this last option.

##### Let's explore built-in data sets:
R and its various packages come with many built-in data sets The data() function lists them. Entering a function name without brackets lists its content, in this case the source code of the function "data".

```{r}
data() # list data sets
cars # let's look at the content of the cars data set

# more information about the data set:
??cars

plot(cars) # plots the data set using the generic plot function
```

We can investigate the structure of the data set. You can access the same information in the Environment tab of the GUI.

```{r}
str(cars) # function to display the structure of the data set
```

The data set is a "data frame", a rectangular composite data structure, similar to an excel table. It consists of two variables, in this case vectors, with each 50 elements. We can look at them individually:

```{r}
cars$speed
cars$dist
```

We can also generate summary statistics of the data frame or each of its variables. For a numerical vector, summary() gives minimum, maximum, mean, median, and the 1st and 2nd quartiles. we can also call functions such as mean(), median() on the variables.

```{r}
summary(cars)
mean(cars$speed)
median(cars$dist)
```

##### Building vectors and data frames

Copy the cars data set into a new data frame, which we will then manipulate. We use the assignment operator <-
We can also make a new variable, here as a result of a calculation.

```{r}
mycars <- cars
mycars$speedratio <- mycars$speed / mycars$dist

# introducing the generic histogram function
hist(mycars$speedratio)

# subsetting / filtering a data frame
mycars[1:5,] # only the first five rows
mycars[,2] # only the second column (here, speed variable)
mycars[5:10,1] # the values of the fifth to tenth row of the first column
mycars[mycars$speedratio<2.5,] #only rows where the speedratio variable is smaller than 2.5

```

#### Let's build a data frame from scratch: 

```{r}
a <- 5 # assign the value five to the variable a
a <- c(1,2,6,8,3) # assign a vector with five elements to variable a. c() - combine values function.
a[1:3] # report first three elements of vector a
b <- c(3,1,5,1,6) # assign a second vector to variable b

c <- data.frame(a,b) # create a data frame c from vectors a and b.
c <- data.frame(first = a, second = b) # same as before, but give new variables different names

c$idx <- seq(1:5) # add a sequence from 1 to 5 as index

c$rnd <- runif(5,min = 1, max = 10) # add a column of five random numbers between 1 and 10

plot(c) # based on the data set, plot generates a scatterplot matrix
```

#### Load an external data set:

```{r}
quake <- read.csv(file = "./data/earthquake.csv", header = T, sep = ",")

str(quake) # note the different non-numerical variables, missing values etc.
summary(quake)

# how are magnitude and depth correlated?
plot(quake$mag,quake$depth)
```

