---
title: "Correlation"
output: 
  html_notebook: 
    theme: simplex
---
Standard chart for examining correlation is the scatter plot. Other elements provide support:

- The ellipse abstracts correlation direction and strength
- The regression line indicates the slope of the correlation coefficient.
```{r}
# Old Faithful Geyser Data: Waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.
plot(faithful)

ggplot(faithful,aes(eruptions,waiting)) + 
  geom_point(alpha = 0.25) +
  stat_ellipse() +
  stat_smooth(method =  "lm", size=0.5, se = F, color = "black")

```


```{r}
# corellation coefficient - 0.9 is quite high.
r <- cor(faithful$waiting, faithful$eruptions)
r
# degree of determination - the variance in eruptions explains 81% of the variance in waiting and vice versa.
r^2

# test of significance - a p < 0.05 is often used to support the alternative h that r is not 0
c <- cor.test(faithful$waiting,faithful$eruptions)
c
c$p.value #only call p-value from result

```

Let's calculate manually.

As a standardized number, correlation requires standardized scores:
The sum of products of the paired z-scores divided by the number of observation pairs.
```{r}
# basic variables
e1 <- faithful$waiting
e2 <- faithful$eruptions
n  <- length(e1) # or e2
x1 <- mean(e1) # 
x2 <- mean(e2)

# for comparison: remember variance (square of standard deviation) 
v <- sum((e1-x1)^2)/ (n-1) # equivalent to var(e1)
sqrt(v) #standard deviation, equivalent to sd(e1)

# covariance is the same, but using the product of 2 variables instead of 1
cv <- sum((e1-x1)*(e2-x2))/ (n-1) # equivalent to cov(e1,e2)

# correlation (pearson) covariance divided by product of standard deviations
r <- cv / (sd(e1)*sd(e2))
r
# or, slightly rearranged:
sum((e1-x1)/sd(e1) * (e2-x2)/sd(e2)) / (n-1)

# manually calculating t-value of the correlation
# equiv. to cor.test(e1,e2)$parameter
t <- r*sqrt((n-2)/(1-r^2))
t
# rule of thumb,  t > 2 is usually significant at 95% level
1-pt(t, df=n-1) # or use lower.tail = F argument
```


For data sets with multiple continuous variables, a scatterplot matrix allows comparison of correlations between all combinations of the variables.

```{r}
# the Fisher iris data set of flower characteristics
plot(iris)

# ggplot does not have a scatterplot matrix, but there are libraries
library(GGally)
ggpairs(iris, aes(color=Species))
```

