---
title: "Time and Date in R"
output:
  html_notebook:
    theme: simplex
  html_document:
    df_print: paged
---
A detailed tutorial of working with time and date data can be found [here](http://www.noamross.net/blog/2014/2/10/using-times-and-dates-in-r---presentation-code.html).
We will work with a json data set of food inspections in Boston. See more [here](https://data.boston.gov/dataset/food-establishment-inspections).

```{r}
# Library to load json files
#install.packages("jsonlite")
library(jsonlite)

url <- "http://data.boston.gov/datastore/odata3.0/4582bec6-2b4f-4f9e-bc55-cbaa73117f4c?$top=150&$format=json"
food <- fromJSON(url)

#inspect the loaded file
head(food)
# the data frame is in the $value list item - that is all we need now.
food <- food$value
```

```{r}
# here we have a date field (date and time of issue) in internet ISO format
food$ISSDTTM[1:5]

# we create a new variable keeping only the date, disregarding time
# in this case, as.Date automatically recognizes the format
food$date <- as.Date(food$ISSDTTM)
food$date[1:5]

# if necessary we could also specify the format manually, for example
as.Date("04/23/2015", format = "%m/%d/%Y")
```


#### Calculation with time
```{r}
# how many inspections on weekdays, months, quarters?
weekdays(food$date)[1:15]
months(food$date)[1:15]
quarters(food$date)[1:15]

#plot inspections per month 
library(ggplot2)
qplot(months(food$date))

# result per weekdays
ggplot(data=food, aes(x=weekdays(date), fill=RESULT)) + geom_bar() 

# date distribution
qplot(food$date, binwidth=15)

# is the first date later than the second?
food$date[1] > food$date[21]

#how many days between the two?
food$date[1] - food$date[21]

difftime(food$date[1], food$date[21], units = 'hours')

# print only years
format(food$date, "%Y")[1:5]
# print only years by numbers
as.numeric(format(food$date, "%Y"))[1:5]
# list only those dates later than 2010
food$date[format(food$date, "%Y") > "2010"]
```

### Posixlt and Posixct
There are two basic classes of date/times. Class "POSIXct" represents the (signed) number of seconds since the beginning of 1970 (in the UTC time zone) as a numeric vector. Class "POSIXlt" is a named list of vectors representing seconds, minutes, hours, days, years etc.


```{r}
#posix is a format that combines date and time 
as.POSIXlt(food$ISSDTTM)[1:5]
tm <- as.POSIXlt("2013-07-24 23:55:26")
str(tm)

# the units are seconds. substract one hour (1000 sec)
as.numeric(tm)
tm
tm - 26

# POSIXlt stores date and time as a list - you can print it via
unlist(tm)

# and access components
tm$mon
tm$hour
tm$min
tm$sec

# years are calculated as difference from 1900
tm$year

# POSIXct is an alternative version, that just stores the number of seconds
tm <- as.POSIXct("2013-07-24 23:55:26")

# this would give us now an error - no longer work
# tm$hour

strptime( tm, format = "%Y-%m-%d")
```
#### Working with Lubridate

The lubridate package was designd to make working with Posix dates and times more convenient.

```{r}
# these operations can be made more convenient with the lubridate package
#install.packages("lubridate")
library(lubridate)

#learn how to use the package functions by reading the vignette
#browseVignettes("lubridate")

#this gives us months, days, years as numbers. 
month(food$date)
day(food$date)
year(food$date) > 2010

# also correctly resolves unix years
year(tm)

# show a histogram only starting the year 2009
f10 <- food[year(food$date)>2009,]
ggplot(f10,aes(x=date)) + geom_histogram()
```


