---
title: "California Food Affordability"
output: html_notebook
---

##This is a dataset from California Health and Human Services about food affordability in Californian female-headed households with children. 

From the summary:

"This table contains data on the average cost of a market basket of nutritious food items relative to income for female-headed households with children, for California, its regions, counties, and cities/towns. The ratio uses data from the U.S. Department of Agriculture and the U.S. Census Bureau."

Link to the data:
https://data.chhs.ca.gov/dataset/food-affordability-2006-2010

First, I import the data and load ggplot.

```{r}
food_data <- read.csv("https://data.chhs.ca.gov/dataset/0114f5bb-4975-419d-95d9-5f0179a8de06/resource/916e2a2e-383b-4af5-9f5b-310500961cb5/download/food-affordability-2006-2010.csv")
library(tidyverse)
install.packages("ggrepel")
library(ggrepel)
str(food_data)
```

The data contains food affordability for all the cities and towns in California. Each city is broken down into eight different races/ethnicities, and a total category. Each race/ethnicity in each city has a median income. There is also a column for the cost of year for food for each city. The affordability ratio represents the median income divided by the cost of food. The other columns in the dataset represent other food affordability ratios which I wasn't able to figure out, and weren't in any documentation. There is also an average family size column for each city.

I tried really hard to filter using vanilla R and I just couldn't figure it out. So I'm using filter() from dplyr (tidyverse).
I'm filtering out the first rows in the dataset, which give me the average results for the whole state of California. I'm also filtering out the "Total" value in the race/ethncity variable, because that is an average as well.

```{r}
filtered_data <- filter(food_data, geoname != "California" & race_eth_name != "Total")
```

Now, I will find some summary statistics for various variables. I don't know how to write functions in R so this will be very inefficient :(

##Summary Statistics of Race/Ethnicities

First, I filter for the data of four races/ethnicities. Then, I calculate the mean of each race's median income and put it in a variable. Then, I make a vector with the labels. Then I combine both vectors into a dataframe.

```{r}
AfricanAm <- filter(filtered_data, race_eth_name == "AfricanAm")
Asian <- filter(filtered_data, race_eth_name == "Asian")
White <- filter(filtered_data, race_eth_name == "White")
Latino <- filter(filtered_data, race_eth_name == "Latino")

AfricanAmMean <- mean(AfricanAm$median_income, na.rm = TRUE)
AsianMean <- mean(Asian$median_income, na.rm = TRUE)
WhiteMean <- mean(White$median_income, na.rm = TRUE)
LatinoMean <- mean(Latino$median_income, na.rm = TRUE)

meansbyRace <- c(AfricanAmMean, AsianMean, WhiteMean, LatinoMean)
labels <- c("African-American", "Asian", "White", "Latino")

labeledMeans <- data.frame(labels, meansbyRace)
labeledMeans <- data.frame(Race_Ethnicity = labels, Average_Median_Household_Income = meansbyRace)
```
Then, I draw the chart.
```{r}
ggplot() +
  geom_col(data=labeledMeans, aes(x=Race_Ethnicity, y=Average_Median_Household_Income)) +
  scale_y_continuous(labels = scales::dollar) +
  xlab("Race") +
  ylab("Median Household Income") +
  ggtitle("Income Data of California Households Headed by a Female")
```

##Mean Affordability Ratio by City

Now, I want to find the mean affordability ratio for various cities, and plot them on a scatterplot. The lower the affordability ratio, the lower the percentage of income goes to food. The higher the affordability ratio, the higher the percentage of income goes to food.

The California cities I am interested in are Carmel, San Diego, San Francisco, Mountain View, San Luis Obispo, Los Angeles, and Malibu. I find the mean affordability ratio for each city and put it in a dataframe with labels.

```{r}
Carmel <- filter(filtered_data, geoname == "Carmel-by-the-Sea city")
SD <-  filter(filtered_data, geoname == "San Diego")
SF <-  filter(filtered_data, geoname == "San Francisco")
Mountain_View <-  filter(filtered_data, geoname == "Mountain View city")
Slo <-  filter(filtered_data, geoname == "San Luis Obispo")
LA <-  filter(filtered_data, geoname == "Los Angeles")
Malibu <-  filter(filtered_data, geoname == "Malibu city")

carmelMean <- mean(Carmel$affordability_ratio, na.rm=TRUE)
sdMean <- mean(SD$affordability_ratio, na.rm=TRUE)
sfMean <- mean(SF$affordability_ratio, na.rm=TRUE)
mountainViewMean <- mean(Mountain_View$affordability_ratio, na.rm=TRUE)
sloMean <- mean(Slo$affordability_ratio, na.rm=TRUE)
laMean <- mean(LA$affordability_ratio, na.rm=TRUE)
malibuMean <- mean(Malibu$affordability_ratio, na.rm=TRUE)

cityMeans <- c(carmelMean, sdMean, sfMean, mountainViewMean, sloMean, laMean, malibuMean)
cityLabels <-c("Carmel", "San Diego", "San Francisco", "Mountain View", "San Luis Obispo", "Los Angeles", "Malibu")
population <- c(3891, 1470000, 870887, 80447, 47536, 3976000, 12879)

city_df <- data.frame(cityLabels, cityMeans, population)
```
Then, I draw the scatterplot. X is the mean affordability ratio and Y is the population, which I entered into a vector manually.
```{r}
ggplot(city_df, aes(x=cityMeans, y=population)) +
  geom_point() +
  scale_y_continuous(labels = scales::comma, limits = c(0, 4000000)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_label_repel(label=cityLabels, size = 3.5) +
  xlab("Mean Affordability Ratio") +
  ylab("Population") +
  ggtitle("How Much Income Goes To Food In California Cities?")
```

According to this graph, affordability ratio is correlated with population. In other words, as population increases, more income tends to go to food. However, this correlation is definitely misleading because this sample is not representative of California as a whole. The smaller populated cities on this graph (Carmel, Malibu, Mountain View) are relatively wealthy, so it would make sense that less income goes to food in these places. Not all small towns in California are wealthy!


