---
title: "Correlations and T Tests"
output: html_notebook
---
### R analysis by Dan Spector

#### This is a dataset from California Health and Human Services about food affordability in Californian female-headed households with children. 

```{r, echo=TRUE, include=FALSE}
food_data <- read.csv("https://data.chhs.ca.gov/dataset/0114f5bb-4975-419d-95d9-5f0179a8de06/resource/916e2a2e-383b-4af5-9f5b-310500961cb5/download/food-affordability-2006-2010.csv")
library(tidyverse)
str(food_data)
```

```{r}
filtered_data <- filter(food_data, geoname != "California" & race_eth_name != "Total")
```

#### Correlation
First, I will investigate if there is a correlation between the median household income field of African-American and Latino households.
```{r}
afr <- filter(filtered_data, race_eth_name == "AfricanAm")
latino <- filter(filtered_data, race_eth_name =="Latino")

##omitting the NA values
afrFiltered <- na.omit(afr$median_income)
latinoFiltered <- na.omit(latino$median_income)



##these vectors are of different lengths, so I'm only using the first 200 entries for each.
cor(afrFiltered[1:200], latinoFiltered[1:200])
```

The Pearson correlation coefficient is 0.15. Since 0 is no correlation and 1 is total positive correlation, African-American and Latino households seem to have a very weak positive correlation.

Now, I will check if the correlation is statistically significant.

```{r}
cor.test(afrFiltered[1:200], latinoFiltered[1:200])
```
The p-value is under 0.05, so this correlation is statistically significant. The alterative hypothesis suggests that the true population correlation is not equal to 0.

I'm not quite sure what to conclude. I definitely do not think there is a causal relationship between African-American and Latino incomes. Maybe there is a just a very weak correlation between these two groups. Each row represents a city in California, so it might make sense that there might be some very week correlation of the groups in each city.

#### T-Test

The t-test will take the mean from a random sample of 200 values from the cost of food food per year category, and compare it with the population mean of the cost of food per year. Because I am comparing a sample mean with a population mean, I will use the 1-sample t-test.

```{r}

##this is the population, NAs omitted
population <- na.omit(filtered_data$cost_yr)

sample <- sample(population, 200)

##mean of population
popMean <- mean(population)

popMean

##1-sample t-test, mu is mean of the population

t.test(sample, mu = popMean)

```
I reran this code block many times in r studio to simulate a sampling distribution. In most cases, the p-value is well over .05, and the alternative hypothesis (that the population mean is not equal to 7268.952) is only rarely supported. In this particular instance, the p-value is .96, suggesting the null hypotheses is supported and the population mean does equal itself.

For this sample, there is 95 percent confidence that the population mean is between 7038.5 and 7489.8. The population mean is between these values.