---
title: "Boston Fires June 2018"
author: "Dan Spector"
date: "9/17/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###This is a dataset of all fires in Boston in June of 2018.

##Setup
Here, I import the dataset and load tidyverse. Then, I attempt to trim all the whitespace from the neighborhood strings. Finally, summary data.

```{r}
data = read.csv("fire_june_2018.csv", header = TRUE, strip.white = TRUE)
summary(data)
```
##Bar Chart
Next, I build a bar chart showing total number of fires by neighborhood. For some reason, I can't get the whitespace in the x labels to go away :(
```{r}
library(tidyverse)
ggplot(data = data) +
  geom_bar(mapping = aes(x = Neighborhood )) +
   theme(text = element_text(size=8),
        axis.text.x = element_text(angle=90, vjust=0.5),
        axis.ticks.x = element_blank()) +
  ylab("Number of Fires") +
  ggtitle("June 2018 Fires in Boston, MA")
```

##Scatterplot
Now, I try to make a scatterplot showing the zip codes on the x axis and the types of roads on the y axis. I tried to figure out how to make the zip codes discrete, but discrete_scale_x wasn't working for me.
```{r}
ggplot(data = data, aes(x=Zip, y=Street.Type)) + 
  geom_point(mapping = aes(x=Zip, y=Street.Type)) +
  xlab("Zip Code") +
  ylab("Road Type")
```

